---
title: "Rats or Cats which sleeps longer?"
author: "Andrew Martin"
date: "21/08/2017"
output:
  pdf_document: default
  html_document: default
  word_document:
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 

In this analysis, we will address the question of which has the longer average sleeping patterns: rats (Rodentia) or cats (Carnivora). 

# Data

The line has *bold* words and also **italic** words.  

# R commands

bleh  

## mean

bleh

## stdev

bleh

## summary

bleh

The data is contained in the `msleep` dataset in the `tidyverse` package. This is loaded:

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
data("msleep")
```

The data consists of `r nrow(msleep)` Species with `r ncol(msleep)` variables measured on it. 

# Cleaning of data

We are only interested in the variables:

- **order:** which is a categorical nominal variable and gives the Order of the Species. 
- **sleep_total:** which is a quantitative continuous variable and gives the total amount of sleep in hours. 

As well, we are only interested in Species that are in the order Carnivora or Rodentia. The following code cleans the dataset:

```{r}
rats_cats  <- msleep %>% 
  filter(order %in% c("Carnivora", "Rodentia")) %>% 
  select(order, sleep_total)
```

```{r,echo = FALSE}
rats_cats%>%
  group_by(order) %>% 
  summarise(sd=sd(sleep_total))%>%
  knitr::kable(digits = 2)
```

# Basic summary statistics

Table \ref{tab:ss} gives the summary statistics for each group. 

```{r, echo = FALSE}
rats_cats %>% 
  group_by(order) %>% 
  summarise(n = n(), mean = mean(sleep_total)) %>% 
  knitr::kable(caption = "Summary statistics for each Order\\label{tab:ss}")
```

# Figuretoldtomake
$$
T = \frac{\bar{X_1} - \bar{X_2}}{\sqrt[]{\frac{S_1^2}{n_1}+\frac{S_2^2}{n_2}}}\\
cov\left(\sum_{i=1}^n a_iY_i , \sum_{j=1}^n b_j Y_j\right) = \sum_{i=1}^n\sum_{j=1}^n
$$
# Side-by-side boxplots

Figure \ref{fig:boxplot} gives the side-by-side boxplots of sleep total for each order. 

```{r, fig.cap="Side-by-side boxplots of total sleep for each order.\\label{fig:boxplot}", echo = FALSE}
ggplot(rats_cats, aes(x = order, y = sleep_total, fill = order)) + 
  geom_boxplot(show.legend = FALSE)
```

# Two-sample t-test

The sample mean for each group was used. To recap the sample mean for each group is 

$$
\bar{Y}_i = \sum^{n_i}_{i = 1}Y_{ij}, 
$$
where $n_i$ is the number in the $i$th group, and $Y_{ij}$ is the $j$th observation in the $i$th group. 

The appropriate null and alternative hypotheses are

$$
\begin{aligned}
H_0 &: \mu_1 - \mu_2 = 0,\\
H_a &: ?? \ne ??,
\end{aligned}
$$
where ??.

We perform a two-sample t-test with 

```{r}
## Insert correct code here for t.test
```

